Pythran: Bringing OpenMP to a Python Subset

### 2013-02-11 doc/papers/europar2013/paper.tex
Take into account most of RK's and MA's comment.
Just for posterity:

* I do not put weave in the biblio because I only use it as an
  introduction to the concept of C interleaved with Python
* Yes, IronPython having no GIL is disturbing. I did not dig too much
  into that part
* I am pretty confident about my threading implementation. No room to say
  everything in the paper, it's just a short intro. It's in the git if
  you're interested :-)
* Simple use of OpenMP is backward compatible, medium advanced us is
  not. I updated the paper accordingly.
* It is not unfair to compare SLOC for C an

\marginpar{The title is to be changed}
%
\begin{abstract}

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that can
communicate through IPC, using Python built-in object serialization. This
approach is only viable for computation intensive application. Furthermore, the
communication and synchronization overheads are much greater than a light-weight
processes approach.

Although the standard \texttt{threading} module makes it possible to start
several light-weight threads within the same interpreter, this approach is not
relevant for HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
is active at a time in the interpreter. While it makes it possible to have
cooperative threads, say for a GUI, it does not take advantage of multiple
cores.  However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. Its execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. So the GIL contention actually \emph{increases} the execution time.

@@

A constraining (from the performance point of view) aspect of the Python
language is its type system. It implies that each method call is resolved
dynamically, even a simple add operation! It comes at no surprise that many
approaches restrain the Python language to add a static typing overlay.  Also,
only two types of integers (64-bit integers and multi-precision integers) and
one type of floating point type (double-precision floats) are available in
Python, so using a type with the appropriate size may lead to significant
performance boost.

@@

The main issue of these approaches is that they imply to modify, in a more or
less complex way, the original code. They require to learn a new dialect,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code were to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

To be completely backward compatible, it must be possible to run the input code
in an environment that is not aware of the existence of the parallelizing
solution. This principle is not respected by parallel libraries, but code
annotations satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
annotations shine: the original code remains compatible with a compiler that is
not aware of OpenMP.

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python,
while being backward compatible with both the Python language and the
sequential algorithm. It means that:
%
\begin{enumerate}
    \item any Pythran code can be run (sequentially), with no module dependency
        or code change, by any Python interpreter;
    \item parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. Using the \texttt{default(none)} clause is possible
to ensure no variable gets forgotten when building such lists.

@@

Pythran runtime is aware of these three kind of iterators: random access
iterators for standard containers, adaptor iterators for generator expressions
and forward iterators for generators.

@@

Providing a binding to these library in Python does not raise particular
problems, as the signature of these functions only involves integers passed as
parameters or return values, except for the mutex manipulation. In that case an
opaque type is used to represent the native type.

@@

Pythran is a subset of Python designed for scientific computing. It is
implicitly statically typed and supports most Python constructs except those
that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
is currently only partially supported. The associated compiler turns Pythran
code, possibly annotated with OpenMP and a few function type annotations, into
C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internals of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

Pythran runtime is based on the STL. The standard library of C++11, unlike its
C++03 version, is thread-safe, so the data dependencies in the original code
remain explicit.

@@

Pythran handles the problem by refusing recursive types, which makes it possible
to solely rely on reference counting for memory management. It can be
implemented through the thread-safe shared pointer mechanism provided in C++11
standard library.

@@

We then compare Pythran with Cython. The approaches share some similarities:
they both generate code written in a lower level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information than Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They translate explicit fine-grained
parallelism through code annotations or specific python constructs,
respectively.

@@

\end{document}

%%% Local Variables:
%%% mode: latex
%%% ispell-local-dictionary: "american"
%%% TeX-auto-untabify: t
%%% TeX-PDF-mode: t
%%% End:

% vim:spell spelllang=en
% vim:tw=80

>>>>

%
\begin{abstract}

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that can
communicate through IPC, using Python built-in object serialization. This
approach is only viable for computation intensive application, since the
communication and synchronization overheads are much greater than a light-weight
processes approach.

Although the standard \texttt{threading} module makes it possible to start
several light-weight threads within the same interpreter, this approach is not
applicable to HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
is active at a time in the interpreter. While it makes it possible to have
cooperative threads, say for a GUI, it does not take advantage of multiple
cores.  However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. Its execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. GIL contention actually \emph{increases} the execution time.

@@

A constraining (from the performance point of view) aspect of the Python
language is its type system. It implies that each method call is resolved
dynamically, even a simple add operation! It comes at no surprise that many
approaches restrain the Python language to add a static typing overlay.  Also,
only two types of integers (64-bits integers and multi-precision integers) and
one type of floating point type (double-precision floats) are available in
Python, so using a type with the appropriate size may lead to significant
performance boost.

@@

The main issue of these approaches is that they imply to modify, in a
backward-incompatible way, the original code. They require to learn a new
dialect, and the long-term preservation of this investment is not ensured.
Moreover, there is no falling back if the code were to be deployed in an
environment where the parallelizing tool/module is not available. For instance,
once a code has been ported to PyCUDA and transformed into CUDA code embedded in
Python strings, there is no way back and the developer has to maintain two
versions of the algorithm.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into native ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

To be completely backward compatible, it must be possible to run the input code
in an environment that is not aware of the existence of the parallelizing
solution. This principle is not respected by parallel libraries, but code
annotations partially satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
annotations shine: the original code remains mostly compatible with a compiler
that is not aware of OpenMP.

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python,
While being backward compatible with both the Python language, and the
sequential algorithm. It means that:
%
\begin{enumerate}
    \item any Pythran code can be run (sequentially), with no module dependency
        or code change, by any Python interpreter;
    \item parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. %Using the \texttt{default(none)} clause is possible
%to ensure no variable gets forgotten when building such lists.

@@

Pythran runtime is aware of these three kind of iterators and accepts supports
parallel iteration over random access iterators and iterators adapting random
ones.

@@

Providing a binding to these library in Python as an \texttt{omp} module does
not raise particular problems, as the signature of these functions only involves
integers passed as parameters or return values, except for the mutex
manipulation. In that case an opaque type is used to represent the native type.

@@

Pythran is a subset of Python designed for scientific computing. It is
implicitly statically typed and supports most Python constructs except those
that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
is currently only partially supported. User classes are not supported. The
associated compiler turns Pythran code, possibly annotated with OpenMP and a few
function type annotations, into C++ code. Its processing is summarized in
Figure~\ref{fig:pythran-compiler} and relies on type inference to remove most
dynamic behavior, combined with a high-level runtime library,
\texttt{pythonic++}, to allow a one-to-one mapping between Python and C++
constructs. The purpose of this section is not to dig into the internals of the
compiler, but rather to focus on the impact of the parallelism layer, especially
on the runtime library.

@@

Pythran runtime is based on the STL. The standard library of C++11, unlike its
C++03 version, is thread-safe, so no data dependencies are added to the original
code.

@@

Pythran handles the problem by refusing recursive types, which makes it possible
to solely rely on reference counting for memory management. It can be
implemented through the thread-safe shared pointer mechanism provided by the C++11
standard library.

@@

We then compare Pythran with Cython. The approaches share some similarities:
they both generate code written in a lower level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information than Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They translate explicit fine-grained
parallelism through code annotations or specific Python constructs,
respectively.

@@

\end{document}
% vim:spell spelllang=en
% vim:tw=80
